---
title: "Core Functions"
output: 
  html_document:
    number_sections: true
    toc: true
pkgdown:
  as_is: true
vignette: >
  %\VignetteIndexEntry{Core Functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Load example dataset
```{r, eval=FALSE}
# Ensure Giotto Suite is installed.
if(!"Giotto" %in% installed.packages()) {
  pak::pkg_install("drieslab/Giotto")
}

# Ensure the Python environment for Giotto has been installed.
genv_exists <- Giotto::checkGiottoEnvironment()
if(!genv_exists){
  # The following command need only be run once to install the Giotto environment.
  Giotto::installGiottoEnvironment()
}
```

```{r, eval=FALSE}
library(Giotto)
g <- GiottoData::loadGiottoMini("vizgen")

# drop pre-computed information (these steps can be ignored)
data_list <- c(g[["images"]], g[["spatial_info"]], g[["feat_info"]])
g <- giotto()
g <- setGiotto(g, data_list)
```


# Spatial Aggregation
For use with raw subcellular data. Find overlapped feature information and convert to a matrix.
Other examples, starting with different data types can be found in the [object creation](object_creation.html) tutorial.
```{r, eval=FALSE}
# overlap of feature points with polygons
g <- calculateOverlap(g)
g <- overlapToMatrix(g)

# specifically use a polygon set & subset of points
g <- calculateOverlap(g,
    spatial_info = "z0", # use polygons from z0
    feat_subset_column = "global_z",
    feat_subset_ids = 0
)
g <- overlapToMatrix(g)

# overlap of image intensity values with polygons
g <- calculateOverlap(g, 
    image_names = c("dapi_z0", "dapi_z1"),
    name_overlap = "image" # default name would be "protein" for image overlaps
)
g <- overlapToMatrix(g,
    type = "intensity", 
    feat_info = "image"
)

# combine overlaps data from multiple spatial units (intended for z stack data)
# also calculate z1
g <- calculateOverlap(g,
    spatial_info = "z1", # use polygons from z1
    feat_subset_column = "global_z",
    feat_subset_ids = 1
)
g <- overlapToMatrix(g, poly_info = "z1")
g = aggregateStacks(g,
    spat_units = c("z0", "z1"),
    feat_type = "rna",
    new_spat_unit = "aggregate",
    values = "raw"
)
```

# Set default Spatial Unit and Feature Type
```{r, eval = FALSE}
# setting active spatial unit and feature type
activeSpatUnit(g) <- "aggregate"
activeFeatType(g) <- "rna"
```

# Spatial Centroid Calculation

```{r, eval=FALSE}
g <- addSpatialCentroidLocations(g, poly_info = "aggregate")
``` 

# Standard workflow

For more detail, see the [standard workflow](general_workflow.html) vignette
```{r, eval=FALSE}
g <- filterGiotto(g,
    expression_threshold = 1,
    feat_det_in_min_cells = 3,
    min_det_feats_per_cell = 5
)
g <- normalizeGiotto(g)
g <- addStatistics(g)
g <- runPCA(g)
g <- runUMAP(g, dimensions_to_use = 1:20)
g <- createNearestNetwork(g, dimensions_to_use = 1:20)
g <- doLeidenCluster(g)
dimPlot2D(g, cell_color = "leiden_clus")
``` 


# Data Access

## Giotto Object Names and Dimensions
*Available with dev version 0.4.0 of GiottoClass*
```{r, eval = FALSE}
# Feature IDs
rownames(g)
featIDs(g)

# Spatial IDs (cells or spots)
colnames(g)
spatIDs(g)

# both sets of IDs
dimnames(g)

# Select specific sets of IDs
featIDs(g, subset = total_expr >= 500)
spatIDs(g, subset = leiden_clus %in% c(4, 5))
spatIDs(g, subset = leiden_clus %in% c(4, 5), negate = TRUE)

# number of cells and features
ncol(g)
nrow(g)
dim(g)
```

## Metadata
```{r, eval=FALSE}
# feature metadata
fDataDT(g)

# cell metadata
pDataDT(g)

# get specific values from cell metadata
g$leiden_clus

# set values directly into cell metadata
# this is safest when directly using values from another metadata column
# since cell_IDs and their ordering may not as expected.
g$leiden_clus2 <- sprintf("clus_%s", g$leiden_clus)
force(g$leiden_clus2)

# annotate based on another column
g <- annotateGiotto(g, 
    name = "region",
    cluster_column = "leiden_clus",
    annotation_vector = c(
        "1" = "a",
        "2" = "b",
        "3" = "c",
        "4" = "a",
        "5" = "d",
        "6" = "b"
    )
)
spatPlot2D(g, cell_color = "region")

# merge in table of values to metadata
ann <- data.frame(
    cell_ID = spatIDs(g),
    random = rnorm(ncol(g))
)
force(ann)
g <- addCellMetadata(g,
    new_metadata = ann,
    by_column = TRUE,
    column_cell_ID = "cell_ID"
)
pDataDT(g)
```

## Giotto Subobjects
Giotto subobjects carry their own metadata about where they should be in
the object. There are several specific getter and setter functions.
GiottoClass 0.4.0 introduces some common APIs
```{r, eval=FALSE}
# also see ?subset_giotto_subobjects
# get any of: “spatial_info”, “spatial_locs”, “spatial_network”, “feat_info”,
#  “expression”, “cell_metadata”, “feat_metadata”, “spatial_enrichment”, 
#  “nn_network”, “dimension_reduction”, “multiomics”, “images"
#  Output is as a list of objects
g[["expression"]]

# be more specific with object name
a <- g[["expression", "raw"]] # (4 items)
# further filter by requested spatial unit. feat_type can also be used.
b <- g[["expression", "raw", spat_unit = "aggregate"]] # (only 1 item)

# any subobject can be set into the Giotto object using setGiotto
g <- setGiotto(g, b)
g <- setGiotto(g, a) # multiple objects can be set at once. 
# Validity is only checked once all items are added.

# pull out all the subobjects of a giotto object
as.list(g)

# Giotto subobjects are S4 structures. They usually wrap a core data object
# This can be extracted using an empty call to `[`
a[[1]][]
```

## Spatial Subobject Coordinates

```{r, eval=FALSE}
# Accessing terra object spatial values can be done either through as.data.table()
gpoly <- g[["spatial_info", "aggregate"]][[1]]
gpoints <- g[["feat_info", "rna"]][[1]]

data.table::as.data.table(gpoly, geom = "XY")
data.table::as.data.table(gpoints, geom = "XY")
data.table::as.data.table(gpoly, geom = "WKT")
data.table::as.data.table(gpoly, geom = "HEX")

# or XY() (GiottoClass 0.4.0 or greater) which works for all spatial objects
# and retrieves xy(z) coordinates as `matrix` and sets them
sl <- g[["spatial_locs", "raw"]][[1]]

# get
m1 <- XY(sl)
m2 <- XY(gpoly)
m3 <- XY(gpoints)
XY(gpoly[1]) # vertices from first polygon

# set
XY(sl) <- m1
XY(gpoly) <- m2
XY(gpoints) <- m3
```

## spatValues

```{r, eval=FALSE}
# See also ?spatValues
# spatValues can pull data from any of: 
# cell expression, cell metadata, spatial locations, spatial enrichment, 
# dimension reduction, polygon info
# The values will be provided as a data.table with the cell_IDs
spatValues(g, feats = c("leiden_clus", "nr_feats"))
# The first object found will be selected by default.
# It is possible to be more specific by mentioning the name of the object to check
spatValues(g, feats = c("Pdgfra", "Mlc1"), expression_values = "normalized")
# Pulling values from multiple spatial units
spatValues(g, feats = c("Pdgfra", "Mlc1"), spat_unit = c("aggregate", "z0", "z1"))
```


# Subset and Join

## Subsetting objects
GiottoClass 0.4.0 introduces `[` and `subset` generic subsetting
```{r, eval=FALSE}
# also see `subset_giotto`
# subset by i (features) and j (observations)
g[1:200, 1:300]

subset(g, subset = leiden_clus %in% c(4, 5))
subset(g, subset = leiden_clus %in% c(4, 5), negate = TRUE)

# The older subsetGiotto() can also be used.
subsetGiotto(g, cell_ids = spatIDs(g)[1:300], feat_ids = featIDs(g)[1:200])

# subset by spatial locations centroids
subsetGiottoLocs(g, y_max = -4900)

# subset subcellular (used when no aggregate information has been calculated yet)
subsetGiottoLocsSubcellular(g, y_max = -4900) # will not work now
```


## Joining objects

```{r, eval=FALSE}
# join objects (without integration) into a single object so that they can be
# analyzed in a common spatial and expression space.

# This updates the cell_IDs and image names, and adds a new `list_ID` column in cell metadata

# dry run can be turned on to preview the spatial positioning of the objects after joining.
joinGiottoObjects(list(g, g), gobject_names = c("a", "b"), dry_run = TRUE)

j <- joinGiottoObjects(list(g, g), gobject_names = c("a", "b"))
```

For a more information, see the object [splitting and joining](split_join.html) tutorial


## Splitting objects

```{r, eval=FALSE}
# split a Giotto object into a list of multiple smaller ones based on a cell metadata column
splitGiotto(g, "leiden_clus")
```

## Slicing objects

```{r, eval=FALSE}
# slice the giotto object based on spatial unit and feature type
z0 <- sliceGiotto(g, spat_unit = "z0")
```

# Session Info  
```{r, eval=FALSE}
sessionInfo()
```
```
R version 4.4.1 (2024-06-14)
Platform: aarch64-apple-darwin20
Running under: macOS Sonoma 14.4

Matrix products: default
BLAS:   /System/Library/Frameworks/Accelerate.framework/Versions/A/Frameworks/vecLib.framework/Versions/A/libBLAS.dylib 
LAPACK: /Library/Frameworks/R.framework/Versions/4.4-arm64/Resources/lib/libRlapack.dylib;  LAPACK version 3.12.0

locale:
[1] en_US.UTF-8/en_US.UTF-8/en_US.UTF-8/C/en_US.UTF-8/en_US.UTF-8

time zone: America/New_York
tzcode source: internal

attached base packages:
[1] stats     graphics  grDevices utils     datasets  methods   base     

other attached packages:
[1] Giotto_4.1.3      GiottoClass_0.4.0

loaded via a namespace (and not attached):
  [1] colorRamp2_0.1.0            DBI_1.2.3                   deldir_2.0-4               
  [4] rlang_1.1.4                 magrittr_2.0.3              GiottoUtils_0.2.0          
  [7] matrixStats_1.4.1           e1071_1.7-14                compiler_4.4.1             
 [10] png_0.1-8                   vctrs_0.6.5                 pkgconfig_2.0.3            
 [13] SpatialExperiment_1.14.0    crayon_1.5.3                fastmap_1.2.0              
 [16] backports_1.5.0             magick_2.8.4                XVector_0.44.0             
 [19] labeling_0.4.3              utf8_1.2.4                  rmarkdown_2.28             
 [22] UCSC.utils_1.0.0            purrr_1.0.2                 xfun_0.47                  
 [25] beachmat_2.20.0             zlibbioc_1.50.0             GenomeInfoDb_1.40.0        
 [28] jsonlite_1.8.9              DelayedArray_0.30.0         BiocParallel_1.38.0        
 [31] terra_1.7-78                irlba_2.3.5.1               parallel_4.4.1             
 [34] R6_2.5.1                    RColorBrewer_1.1-3          reticulate_1.39.0          
 [37] GenomicRanges_1.56.0        scattermore_1.2             Rcpp_1.0.13                
 [40] SummarizedExperiment_1.34.0 knitr_1.48                  R.utils_2.12.3             
 [43] FNN_1.1.4.1                 IRanges_2.38.0              Matrix_1.7-0               
 [46] igraph_2.0.3                tidyselect_1.2.1            rstudioapi_0.16.0          
 [49] abind_1.4-8                 codetools_0.2-20            lattice_0.22-6             
 [52] tibble_3.2.1                Biobase_2.64.0              withr_3.0.1                
 [55] evaluate_1.0.0              sf_1.0-16                   units_0.8-5                
 [58] proxy_0.4-27                exactextractr_0.10.0        pillar_1.9.0               
 [61] MatrixGenerics_1.16.0       KernSmooth_2.23-24          checkmate_2.3.2            
 [64] stats4_4.4.1                plotly_4.10.4               generics_0.1.3             
 [67] dbscan_1.2-0                sp_2.1-4                    S4Vectors_0.42.0           
 [70] ggplot2_3.5.1               munsell_0.5.1               scales_1.3.0               
 [73] GiottoData_0.2.15           gtools_3.9.5                class_7.3-22               
 [76] glue_1.8.0                  lazyeval_0.2.2              tools_4.4.1                
 [79] GiottoVisuals_0.2.5         data.table_1.16.0           ScaledMatrix_1.12.0        
 [82] cowplot_1.1.3               grid_4.4.1                  tidyr_1.3.1                
 [85] colorspace_2.1-1            SingleCellExperiment_1.26.0 GenomeInfoDbData_1.2.12    
 [88] raster_3.6-26               BiocSingular_1.20.0         rsvd_1.0.5                 
 [91] cli_3.6.3                   fansi_1.0.6                 S4Arrays_1.4.0             
 [94] viridisLite_0.4.2           dplyr_1.1.4                 uwot_0.2.2                 
 [97] gtable_0.3.5                R.methodsS3_1.8.2           digest_0.6.37              
[100] BiocGenerics_0.50.0         classInt_0.4-10             SparseArray_1.4.1          
[103] ggrepel_0.9.6               farver_2.1.2                rjson_0.2.21               
[106] htmlwidgets_1.6.4           htmltools_0.5.8.1           R.oo_1.26.0                
[109] lifecycle_1.0.4             httr_1.4.7     
```
