---
title: "Interactive viewing with GiottoLens"
tags:
  packages: [GiottoDisk, GiottoLens]
  size: full
  platform: Atera
  panel: "5,000+"
  species: human
  tissue: breast
  condition: cancer
  modality: RNA
  resolution: single cell
output:
  html_document:
    number_sections: true
    toc: true
pkgdown:
  as_is: true
vignette: >
  %\VignetteIndexEntry{Interactive viewing with GiottoLens}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

# Overview

GiottoLens is an interactive browser viewer for GiottoDisk-backed Giotto
objects. It draws every transcript and every cell polygon of a whole section at
interactive frame rates, and stays connected to your R session: metadata is
pushed from R into the viewer, and cell selections and drawn regions come back
from the viewer into R.

This tutorial continues from the
[Atera Human Breast Cancer](atera_human_breast_cancer.html) example, using the
clustered object saved at the end of it. It covers:

- launching the viewer on a backed Giotto object
- coloring cells by clusters and other metadata, with a linked UMAP panel
- selecting cells in the viewer and using the selection in R
- drawing regions and bringing them back as `SpatVector` polygons
- pushing selections and polygons from R into the viewer

# How it works

`launch_lens()` starts a small local web server from R and opens the viewer in
your browser or the IDE viewer pane. The viewer reads the transcript and
polygon parquet files that GiottoDisk already wrote to the project directory,
so launching does not copy or convert the data. All rendering and querying
happens in the browser (deck.gl for drawing, DuckDB WASM for queries), and the
viewer bundles everything it needs, so it runs offline and on HPC nodes.

Because the viewer reads GiottoDisk's on-disk stores directly, it needs a
**backed** Giotto object. An in-memory object has no files for it to read.

# Install

```{r}
# GiottoDisk, if it is not installed yet
GiottoUtils::suite_install("GiottoDisk")

# GiottoLens
remotes::install_github("giotto-suite/GiottoLens")
```

# Setup

Load the clustered object saved at the end of the
[Atera example](atera_human_breast_cancer.html).

```{r}
library(arrow) # this must be libraried before terra or GiottoClass
library(Giotto)
library(GiottoDisk)
library(GiottoLens)

project_dir <- "path/to/atera_project"

g <- loadGiotto(project_dir, name = "clustered")
```

# Launch the viewer

```{r}
viewer <- launch_lens(g)
viewer
```

```
LensViewer
  url:          http://localhost:3000
  connected:    TRUE
  selection:    0 cell(s)
  annotations:  0 polygon(s)
```

By default the viewer shows the `"rna"` transcripts and the `"cell"`
polygons. Choose others with `feat_type =` and `spat_unit =`. Use `port =` if
3000 is already taken.

`launch_lens()` returns straight away. The server keeps running in the
background, so you can keep working in the R console while the viewer is open.

<!-- figure: the viewer at launch, whole section with density overview -->

## Running on a remote machine

On an HPC node or a remote server, launch without opening a browser and
forward the port to your own machine:

```{r}
viewer <- launch_lens(g, open = FALSE, port = 3000)
```

```sh
# on your own machine
ssh -L 3000:localhost:3000 user@remote-host
```

Then open `http://localhost:3000/lens/` in your local browser.

# Finding your way around

The layers switch on as you zoom in:

| Layer | Shown |
|---|---|
| Density overview | at low zoom; follows the gene selection |
| Cell centroids | at all zooms, once polygons load |
| Cell outlines | past the polygon zoom threshold |
| Transcripts | past the transcript zoom threshold, subsampled until fully zoomed in |

Every layer has its own visibility toggle and opacity slider in the sidebar.
Pick genes in the sidebar to show only their transcripts, and the density
overview follows the same selection.

Keyboard shortcuts:

| Key | Action |
|---|---|
| `Space` | switch to pan |
| `Esc` | cancel the current drawing and return to pan |
| `c` | clear selection filters |
| `x` | clear the ruler and transcript lasso |
| `[` / `]` | shrink / grow the brush |
| `Delete` | delete the selected annotation |

The camera button in the toolbar saves the current view as a PNG.

<iframe style="width: 100%; aspect-ratio: 16 / 9;" src="https://www.youtube.com/embed/_k0zJj4bqZo" title="GiottoLens - Navigation" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture; web-share" referrerpolicy="strict-origin-when-cross-origin" allowfullscreen></iframe>

<!-- figure: zoomed-in view with transcripts and cell outlines -->

# Color cells by metadata

Push cell metadata from R to color the polygons by it. Adding a dimension
reduction also fills the linked UMAP panel.

This can be done directly from the gobject or from the output of `prep_meta`
which lets you preview and modify the metadata before it is sent.

Name each column with the type the viewer should treat it as:
`"categorical"` or `"continuous"`.

```{r}
meta <- prep_meta(g,
    list(
        leiden_clus = "categorical",
        total_expr  = "continuous",
        nr_feats    = "continuous"
    ),
    dimreds = "umap"
)
head(meta)
```

```
     cell_ID leiden_clus total_expr nr_feats     umap_1      umap_2
1 aaaajgij-1           1       5460     3410 -1.3395577 -0.59741985
2 aaaandia-1           1       6471     3746 -1.4628397 -0.57696640
3 aaabalki-1           2       1609     1377  1.4673266 -2.73528539
4 aaabchln-1           2        809      742  2.3197617 -2.44874393
5 aaaccnnp-1           1       2855     2000 -0.5828114 -0.38163435
6 aaacdfgp-1           1       4108     2571 -1.5705419 -0.07051479
```

```{r}
viewer$push_metadata(meta)
```

The color-by selector in the sidebar now lists the pushed columns.
Categorical columns get a legend, and continuous columns get a gradient.
Columns given without a type keep their R type: factors, characters and
integers are categorical, other numbers are continuous. The UMAP panel uses
the same colors as the spatial view.

Hover over a cell, as a centroid or an outline, to see its metadata in a
pop-up. The column you are coloring by is listed first, then the other pushed
columns, up to 12 in all. Before any metadata is pushed, the pop-up shows only
the cell ID. It is hidden while you lasso, brush or draw.

Column names are cleaned up to letters, digits and underscores on the way in,
so `umap.projection_1` shows up as `umap_projection_1`.

<!-- figure: cells colored by leiden_clus, with the UMAP panel open -->

# Select cells

<iframe style="width: 100%; aspect-ratio: 16 / 9;" src="https://www.youtube.com/embed/x2jEaNDZDSE" title="GiottoLens - Selection and Annotation" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture; web-share" referrerpolicy="strict-origin-when-cross-origin" allowfullscreen></iframe>

The select tools in the toolbar:

- **Lasso**: draw around cells
- **Brush**: paint cells in or out; `[` and `]` resize it
- **Invert**: flip the current selection
- **Select by metadata**: pick cells by a column value

Selections are linked across panels. Lasso on the UMAP and the matching cells
light up in the tissue, and the reverse. With **cross-filter** mode on, a
spatial selection and a UMAP or histogram selection intersect, so you can ask
for "cluster 3, but only inside this duct".

<!-- figure: a lasso selection, highlighted in both the tissue and the UMAP -->

## Use the selection in R

The current selection is always available as `viewer$selection`, a character
vector of cell IDs that updates as you select:

```{r}
ids <- viewer$selection
length(ids)
```

From there it is a regular set of cell IDs. For example, record it as a
metadata column and push it back so it can be colored by:

```{r}
g <- addCellMetadata(g,
    new_metadata = data.frame(
        cell_ID = spatIDs(g),
        lens_selection = ifelse(spatIDs(g) %in% ids, "selected", "other")
    ),
    by_column = TRUE,
    column_cell_ID = "cell_ID"
)

meta <- prep_meta(g,
    list(
        leiden_clus    = "categorical",
        total_expr     = "continuous",
        nr_feats       = "continuous",
        lens_selection = "categorical"
    ),
    dimreds = "umap"
)

viewer$push_metadata(meta)
```

Or compare the selected cells against everything else:

```{r}
markers <- findMarkers(g,
    method = "scran",
    expression_values = "normalized",
    cluster_column = "lens_selection",
    group_1 = "selected",
    group_2 = "other"
)
```

## Push a selection from R

`set_selection()` goes the other way, highlighting cells chosen in R. This
turns off cross-filter mode so the pushed cells show exactly as given.

```{r}
cx <- pDataDT(g)
viewer$set_selection(cx[leiden_clus == 3]$cell_ID)

# an empty vector clears it
viewer$set_selection(character(0))
```

# Draw regions

The draw tools (polygon, rectangle and freehand) make named, persistent
annotations, listed in the sidebar. Each is returned to R in
`viewer$annotations` as GeoJSON:

```{r}
annots <- viewer$annotations
```

Convert them to a `SpatVector` for use with terra or Giotto's spatial tools.
Coordinates are in the same space as the Giotto object, so no transform is
needed:

```{r}
regions <- lapply(annots, function(a) {
    v <- terra::vect(a$geojson, crs = "local")
    v$name <- a$name
    v
})
regions <- do.call(rbind, regions)

plot(regions)
```

<!-- figure: drawn annotations in the viewer -->

## Push polygons from R

`push_polygons()` shows any polygons from R as a semi-transparent layer,
above the density overview and below the cell outlines. It accepts a
`SpatVector`, an `sf` object, or WKT strings, and is useful for spatial
domains, tissue regions or anything else in the object's coordinate space.

```{r}
viewer$push_polygons(regions, names = regions$name)
```

# Measure

The measure tools:

- **Ruler**: click a start and an end point; the distance is shown in μm or mm
- **Transcript lasso**: draw a region to list its 50 most abundant genes. This
  counts every transcript in the region, not the subsample drawn at lower zoom
  levels.

Measurements stay on screen when you switch tools until you clear them with
the measure group's clear button or `x`.

<!-- figure: transcript lasso with the top-genes panel -->

# Close the viewer

```{r}
viewer$close()
```

The server also stops when the R session ends.

# Session info

```{r}
sessionInfo()
```
