---
title: "Create and change Giotto instructions"
output: 
  html_document:
    number_sections: true
    toc: true
pkgdown:
  as_is: true
vignette: >
  %\VignetteIndexEntry{Create and change Giotto instructions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

The Giotto instructions slot contains values that homogenize and facilitate the behavior of the Giotto object, such as whether the plots should be stored locally, shown, or returned, where to save the plots, what python path to use, etc. 

If you don't provide specific instructions, some default values will be automatically added. 

# Create new instructions

Create new instructions and save them in a variable to pass them to the Giotto object later.

```{r, eval=FALSE}
results_folder <- "path/to/results"
python_path <- NULL

instructions <- createGiottoInstructions(save_dir = results_folder,
                                         save_plot = TRUE,
                                         show_plot = FALSE,
                                         return_plot = FALSE,
                                         python_path = python_path)

createGiottoObject(expression = my_expression,
                   spatial_locs = my_spatial_locs,
                   instructions = instructions)
```

# Retrieve instructions

## Setup and load example dataset

```{r, eval=FALSE}
# Ensure Giotto Suite is installed
if(!"Giotto" %in% installed.packages()) {
  pak::pkg_install("drieslab/Giotto")
}

# Ensure Giotto Data is installed
if(!"GiottoData" %in% installed.packages()) {
  pak::pkg_install("drieslab/GiottoData")
}

library(Giotto)

# Ensure the Python environment for Giotto has been installed
genv_exists <- checkGiottoEnvironment()

if(!genv_exists){
  # The following command need only be run once to install the Giotto environment
  installGiottoEnvironment()
}
```

```{r, eval=FALSE}
# load the object
g <- GiottoData::loadGiottoMini("visium")
```

## Show current instructions

```{r, eval=FALSE}
instructions(g)
```

# Update the instructions 

## Update one instruction value

```{r, eval=FALSE}
instructions(g, "return_plot") <- TRUE
```

## Update multiple values

```{r, eval=FALSE}
instructions(g, c("return_plot", "save_plot")) <- c(TRUE, TRUE)
```

# Session Info

```{r, eval=FALSE}
sessionInfo()
```

```{r, eval=FALSE}
R version 4.4.2 (2024-10-31)
Platform: x86_64-apple-darwin20
Running under: macOS Sequoia 15.0.1

Matrix products: default
BLAS:   /System/Library/Frameworks/Accelerate.framework/Versions/A/Frameworks/vecLib.framework/Versions/A/libBLAS.dylib 
LAPACK: /Library/Frameworks/R.framework/Versions/4.4-x86_64/Resources/lib/libRlapack.dylib;  LAPACK version 3.12.0

locale:
[1] en_US.UTF-8/en_US.UTF-8/en_US.UTF-8/C/en_US.UTF-8/en_US.UTF-8

time zone: America/New_York
tzcode source: internal

attached base packages:
[1] stats     graphics  grDevices utils     datasets  methods   base     

other attached packages:
[1] Giotto_4.1.4      GiottoClass_0.4.3

loaded via a namespace (and not attached):
 [1] tidyselect_1.2.1            viridisLite_0.4.2          
 [3] dplyr_1.1.4                 GiottoVisuals_0.2.7        
 [5] fastmap_1.2.0               SingleCellExperiment_1.26.0
 [7] lazyeval_0.2.2              digest_0.6.37              
 [9] lifecycle_1.0.4             terra_1.7-78               
[11] magrittr_2.0.3              compiler_4.4.2             
[13] rlang_1.1.4                 tools_4.4.2                
[15] igraph_2.0.3                utf8_1.2.4                 
[17] yaml_2.3.10                 data.table_1.16.0          
[19] knitr_1.48                  S4Arrays_1.4.1             
[21] htmlwidgets_1.6.4           reticulate_1.39.0          
[23] DelayedArray_0.30.1         abind_1.4-8                
[25] withr_3.0.1                 purrr_1.0.2                
[27] BiocGenerics_0.50.0         grid_4.4.2                 
[29] stats4_4.4.2                fansi_1.0.6                
[31] colorspace_2.1-1            ggplot2_3.5.1              
[33] scales_1.3.0                gtools_3.9.5               
[35] SummarizedExperiment_1.34.0 cli_3.6.3                  
[37] rmarkdown_2.28              crayon_1.5.3               
[39] generics_0.1.3              rstudioapi_0.16.0          
[41] httr_1.4.7                  rjson_0.2.23               
[43] zlibbioc_1.50.0             parallel_4.4.2             
[45] XVector_0.44.0              matrixStats_1.4.1          
[47] vctrs_0.6.5                 Matrix_1.7-1               
[49] jsonlite_1.8.9              GiottoData_0.2.15          
[51] IRanges_2.38.1              S4Vectors_0.42.1           
[53] ggrepel_0.9.6               scattermore_1.2            
[55] magick_2.8.5                GiottoUtils_0.2.0          
[57] plotly_4.10.4               tidyr_1.3.1                
[59] pak_0.8.0                   glue_1.8.0                 
[61] codetools_0.2-20            cowplot_1.1.3              
[63] gtable_0.3.5                GenomeInfoDb_1.40.1        
[65] GenomicRanges_1.56.1        UCSC.utils_1.0.0           
[67] munsell_0.5.1               tibble_3.2.1               
[69] pillar_1.9.0                htmltools_0.5.8.1          
[71] GenomeInfoDbData_1.2.12     R6_2.5.1                   
[73] evaluate_1.0.0              lattice_0.22-6             
[75] Biobase_2.64.0              png_0.1-8                  
[77] backports_1.5.0             SpatialExperiment_1.14.0   
[79] Rcpp_1.0.13                 SparseArray_1.4.8          
[81] checkmate_2.3.2             colorRamp2_0.1.0           
[83] xfun_0.47                   MatrixGenerics_1.16.0      
[85] pkgconfig_2.0.3 
```
