---
title: "Create and change Giotto instructions"
output: 
  html_document:
    number_sections: false
    toc: true
pkgdown:
  as_is: true
vignette: >
  %\VignetteIndexEntry{Create and change Giotto instructions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, eval=FALSE}
# Ensure Giotto Suite is installed
if(!"Giotto" %in% installed.packages()) {
  pak::pkg_install("drieslab/Giotto")
}
library(Giotto)
```

# Giotto Instructions

Many {Giotto} analyses start with manually defining a set of instructions to use.
{Giotto} instructions are used to apply settings to `giotto` object behavior at the
project level. Each `giotto` object needs a set of instructions information
that can either be manually set or generated with defaults when the `giotto`
object is first created via either `createGiotto*()` helpers or the S4 class
constructor `giotto()`.
Once added, these instructions are stored in the `giotto` object `@instructions`
slot.

These instruction sets can be created via either the constructor
```{r, eval=FALSE}
createGiottoInstructions()
```

or the generic accessor/replacement function `instructions()` when no `giotto`
object is provided.
```{r, eval=FALSE}
# named params can be provided (same as those in createGiottoInstructions())
instructions()
```

These instruction sets are of S3 class `giottoInstructions`, extending `list`.

Specific information about all behaviors that can be set in these instruction
sets can be found in the `?GiottoClass::createGiottoInstructions` documentation.

# Commonly Used Instructions

The most commonly used settings are:

* `save_dir` - directory to save plot results to
* save_plot - logical. Whether plotting functions should save outputs to the
`save_dir`
* `show_plot` - logical. Whether plotting functions should display the plot
* `return_plot` - logical. Whether a plot object (for example a {ggplot2} `gg` 
object) should be returned from the plotting function.
* `python_path` - filepath to a python binary to use with {Giotto}.

A commonly used set of instructions:

```{r, eval=FALSE}
results_folder <- "path/to/results"
python_path <- NULL # NULL will let {Giotto} autodetect

intrs <- createGiottoInstructions(
    save_dir = results_folder,
    save_plot = TRUE,
    show_plot = FALSE,
    return_plot = FALSE,
    python_path = python_path
)
```

This establishes the location defined by `results_folder` as a common location
to export all plots created from that `giotto` object as saved images. This can
be helpful for setting up analysis workflows.

`save_plot` is set to `TRUE`, while `show_plot` is set to `FALSE`. This
ensures that plots are saved but not plotted directly to the viewer. Outputting
plots in this manner can save time since directly plotting can be slow.

`return_plot` is set as `FALSE` since it is not necessary in most cases. It
additionally prevents double plotting form occurring when the plotting function
output is not assigned into a variable.

A python path to a specified python binary is also specified here, although
it is usually preferable to set the following global option in the `~/.Rprofile`
when the same particular path is needed every time, and then allow autodetection
to find it. (see next section)

```{r, eval=FALSE}
options("giotto.py_path" = "???")
```

## Python Environment Considerations

The python environment to use via {reticulate} can be set with the `python_path`
param. Providing `NULL` will have {Giotto} autodetect for a python environment
to use (usually the `giotto_env`).

 - See [Giotto python configuration](configuration.html) and/or
 `?GiottoClass::giotto_python` for more info on how to set up a
  {Giotto} miniconda python environment and how autodetection of an environment
  to use works.

Creating a set of instructions will call
`GiottoClass::set_giotto_python_path()` under the hood and activate a python
environment.

**Critically, only one {reticulate} python environment can be activated per session.**
If you would like to avoid activating a python environment,
you can set the following global option before running a `giotto` or
`giottoInstructions` object creation function:

```{r, eval=FALSE}
options("giotto.use_conda" = FALSE)
```



# Manipulating Giotto Instructions

## Applying During `giotto` Object Creation

Giotto instructions can be set or retrieved from a `giotto` object at any time.
They can be initially applied directly at the `giotto` object creation step.
Here we add in the `giottoInstructions` that we created in the previous section.

```{r, eval=FALSE}
createGiottoObject(instructions = instrs) # object creation helper
# or
giotto(instructions = instrs) # class constructor
```

*Of course the actual data can also be added via other params during these
`giotto` object creation steps, they are just omitted for this demonstration*


## Getting instructions

You can extract the full set of instructions in a `giotto` object using the
`instructions()` generic.

```{r, eval=FALSE}
g <- giotto() # will generate and attach a default set of instructions
i <- instructions(g)
force(i)
```

You can also extract specific parameters of that instruction set
```{r, eval=FALSE}
instructions(g, "dpi")
```

This will also work directly on the `giottoInstructions` object
```{r, eval=FALSE}
instructions(i, "dpi")
```

## Setting instructions

Setting of `giottoInstructions` is also done with the `instructions()` generic.

```{r, eval=FALSE}
instructions(g) <- instrs # `instrs` had `return_plot = FALSE`
instructions(g, "return_plot") # this is now FALSE
```

You can also update one or more specific instructions settings at once either
directly on the `giottoInstructions` object or to the `giotto` object.

## Single value updates

```{r, eval=FALSE}
# on the giottoInstructions object
instructions(i, "dpi") <- 600
force(i)
# on the giotto object
instructions(g, "dpi") <- 600
force(instructions(g))
```

## Update multiple value

Multiple value replacement currently only works with the `giotto` object.
```{r, eval=FALSE}
instructions(g, c("plot_format", "height")) <- list("tiff", 15)
force(instructions(g))
```

# Session Info

```{r, eval=FALSE}
sessionInfo()
```

```{r, eval=FALSE}
R version 4.4.1 (2024-06-14)
Platform: aarch64-apple-darwin20
Running under: macOS 15.3.1

Matrix products: default
BLAS:   /System/Library/Frameworks/Accelerate.framework/Versions/A/Frameworks/vecLib.framework/Versions/A/libBLAS.dylib 
LAPACK: /Library/Frameworks/R.framework/Versions/4.4-arm64/Resources/lib/libRlapack.dylib;  LAPACK version 3.12.0

locale:
[1] en_US.UTF-8/en_US.UTF-8/en_US.UTF-8/C/en_US.UTF-8/en_US.UTF-8

time zone: America/New_York
tzcode source: internal

attached base packages:
[1] stats     graphics  grDevices utils     datasets  methods   base     

other attached packages:
[1] Giotto_4.2.1      GiottoClass_0.4.7

loaded via a namespace (and not attached):
  [1] colorRamp2_0.1.0            gridExtra_2.3               remotes_2.5.0              
  [4] rlang_1.1.4                 magrittr_2.0.3              GiottoUtils_0.2.4          
  [7] matrixStats_1.4.1           compiler_4.4.1              png_0.1-8                  
 [10] callr_3.7.6                 vctrs_0.6.5                 pkgconfig_2.0.3            
 [13] SpatialExperiment_1.14.0    crayon_1.5.3                fastmap_1.2.0              
 [16] backports_1.5.0             magick_2.8.5                XVector_0.44.0             
 [19] ggraph_2.2.1                utf8_1.2.4                  rmarkdown_2.29             
 [22] UCSC.utils_1.0.0            ps_1.7.6                    purrr_1.0.2                
 [25] xfun_0.49                   zlibbioc_1.50.0             cachem_1.1.0               
 [28] GenomeInfoDb_1.40.0         jsonlite_1.8.9              DelayedArray_0.30.0        
 [31] tweenr_2.0.3                terra_1.8-29                parallel_4.4.1             
 [34] R6_2.5.1                    reticulate_1.39.0           pkgload_1.3.4              
 [37] GenomicRanges_1.56.0        scattermore_1.2             Rcpp_1.0.13-1              
 [40] SummarizedExperiment_1.34.0 knitr_1.49                  IRanges_2.38.0             
 [43] Matrix_1.7-0                igraph_2.1.1                tidyselect_1.2.1           
 [46] rstudioapi_0.16.0           abind_1.4-8                 yaml_2.3.10                
 [49] viridis_0.6.5               codetools_0.2-20            curl_5.2.3                 
 [52] processx_3.8.4              pkgbuild_1.4.4              lattice_0.22-6             
 [55] tibble_3.2.1                Biobase_2.64.0              withr_3.0.2                
 [58] evaluate_1.0.1              desc_1.4.3                  polyclip_1.10-7            
 [61] pillar_1.9.0                MatrixGenerics_1.16.0       checkmate_2.3.2            
 [64] stats4_4.4.1                plotly_4.10.4               generics_0.1.3             
 [67] S4Vectors_0.42.0            ggplot2_3.5.1               munsell_0.5.1              
 [70] scales_1.3.0                GiottoData_0.2.15           gtools_3.9.5               
 [73] glue_1.8.0                  lazyeval_0.2.2              tools_4.4.1                
 [76] GiottoVisuals_0.2.12        data.table_1.16.2           graphlayouts_1.1.1         
 [79] tidygraph_1.3.1             cowplot_1.1.3               grid_4.4.1                 
 [82] tidyr_1.3.1                 colorspace_2.1-1            SingleCellExperiment_1.26.0
 [85] GenomeInfoDbData_1.2.12     ggforce_0.4.2               cli_3.6.3                  
 [88] dbMatrix_0.0.0.9023         fansi_1.0.6                 S4Arrays_1.4.0             
 [91] viridisLite_0.4.2           dplyr_1.1.4                 gtable_0.3.6               
 [94] digest_0.6.37               BiocGenerics_0.50.0         SparseArray_1.4.1          
 [97] ggrepel_0.9.6               rjson_0.2.21                htmlwidgets_1.6.4          
[100] farver_2.1.2                memoise_2.0.1               htmltools_0.5.8.1          
[103] lifecycle_1.0.4             httr_1.4.7                  MASS_7.3-60.2     
```
