---
title: "Data Normalization and Scaling"
output: 
  html_document:
    number_sections: true
    toc: true
pkgdown:
  as_is: true
vignette: >
  %\VignetteIndexEntry{Affine Image Registration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, eval=FALSE}
# Ensure Giotto Suite is installed.
if(!"Giotto" %in% installed.packages()) {
    remotes::install_github("drieslab/GiottoUtils")
    GiottoUtils::suite_install()
}

# Ensure the Python environment for Giotto has been installed.
genv_exists <- Giotto::checkGiottoEnvironment()
if(!genv_exists){
  # The following command need only be run once to install the Giotto environment.
  Giotto::installGiottoEnvironment()
}
```

# Expression Value Processing Pipeline

{Giotto} `v4.2.1` and {GiottoClass} `v0.4.7` add a composable framework for
data normalization and transformation with 2 main components:

* `processExpression()` - `giotto` object-level function for
**working with the expression values**. Passes to `processData()` after handling
any `giotto` object-level data extraction/processing.
    * `processData()` - lower-level generic that handles specific operations at the
data representation level.
* Factory functions for method-specific **parameter classes** that both define the operation
to perform and also contain all relevant parameters in an editable format. For
the full set of available methods, see `?process_param`
    * `normParam()` - data normalization
    * `scaleParam()` - data scaling
    * `adjustParam()` - matrix adjustments


*This workflow supercedes the original `normalizeGiotto()` function.*

## Single Operations

**Performing a simple log2 transform:**
```{r, eval=FALSE}
library(Giotto)
# load example data to use
g <- GiottoData::loadGiottoMini("visium")

logparam <- normParam("log")
force(logparam) # class of "logNormParam"

g <- processExpression(g,
    expression_values = "raw", # (default = "raw") name of expression values to use
    param = logparam, # parameter object to use
    name = "log" # name to apply to resulting transformed values
)

# extract and display values
force(g[["expression", "log"]])
```

**Changing params in a parameter class:**

Get param settings in a parameter class using the `$` operator. Available params
will tab autocomplete. Param definitions are also documented in the help
pages of specific parameter classes.

See also:

* `?process_param` - central directory of provided processing parameters
* `?norm_log` - help page for log normalization param

```{r, eval=FALSE}
# changing a parameter (available param names will autocomplete)
force(logparam$base)
logparam$base <- 10 # change to base 10 log

g <- processExpression(g,
    param = logparam,
    name = "log10"
)

# extract and display values
force(g[["expression", "log10"]])
```

## Chained Operations

**These operations can additionally be chained together when provided as list:**
```{r, eval=FALSE}
g <- processExpression(g, 
    list(normParam("log"),
         normParam("library")),
    name = "log_lib"
)

# extract and display values
force(g[["expression", "log_lib"]])
```


# Example workflows

## Default Normalization and Scaling

The default normalization and scaling method as implemented by
`normalizeGiotto()` is:

1. library normalization
2. log normalization
3. zscore scaling along features
4. zscore scaling along cells/samples/observations

With `processExpression()`, this can be rewritten as either of the two below:

Using "default" params:

```{r, eval=FALSE}
g <- processExpression(g, 
    normParam(),
    name = "normalized"
)

g <- processExpression(g,
    expression_values = "normalized",
    scaleParam(),
    name = "scaled"
)
```

or written out fully:

```{r, eval=FALSE}
g <- processExpression(g,
    list(
        normParam("library"),
        normParam("log")
    ),
    name = "normalized"
)

g <- processExpression(g,
    expression_values = "normalized",
    list(
        scaleParam("zscore", MARGIN = 1), # along rows (features)
        scaleParam("zscore", MARGIN = 2) # along cols (samples)
    ),
    name = "scaled"
)
```

See also:

* `?norm_log`
* `?norm_library`
* `?scale_zscore`

## Pearson Residual Normalization

A normalization method described in [Lause, 2021](https://doi.org/10.1186/s13059-021-02451-7),
designed for detection of highly variable features and dimension reduction/clustering.
This method includes a z-score-like transformation, so scaling is not recommended after
this. Using these values for differential gene expression analysis is also not
recommended.

```{r, eval=FALSE}
g <- processExpression(g,
    normParam("pearson"),
    name = "pearson"
)
```

See also:

* `?norm_pearson`

## Arcsinh Transformation with Cofactor C

Arcsinh transformation is commonly used with spatial proteomics/intensity type data.
It is suited for data with a wide dynamic range and zero/near-zero values and
preserves detail across multiple orders of magnitude, where a linear or log
transform would distort important features.

A **cofactor C** controls the transition between linear and logarithmic behaviors.
With CODEX or CyCIF, `c = 5` is commonly used. `c = 1` or `5` is commonly used with
IMC.

* Smaller C values (1-3): More aggressive transformation, useful for very wide dynamic ranges
* Larger C values (5-10): More gentle transformation, preserving more of the original data structure

```{r, eval=FALSE}
anorm <- normParam("arcsinh")
anorm$c <- 5 # (this is the default c value)

g <- processExpression(g, list(
    anorm,
    scaleParam("zscore", MARGIN = 2) # zscore across samples
), name = "arcsinh")
```

See also:

* `?norm_arcsinh`
* `?scale_zscore`

## TF-IDF Normalization

TF-IDF (Term Frequency-Inverse Document Frequency) is commonly used in ATAC-seq
data, but is generally helpful in situations where the data is very noisy. This
normalization method pulls on concepts of natural language processing to
effectively focus in on features specific to particular samples/cells.

```{r, eval=FALSE}
g <- processExpression(g, list(
    normParam("tf-idf"),
    normParam("l2") # L2 normalization (Euclidean normalization) is commonly performed afterwards
), name = "tfidf")
```

See also:

* `?norm_tfidf`
* `?norm_l2`

# Adding A Custom Method

Processing methods for this framework are defined for the `processData()`
generic from {GiottoClass} against **parameter class** and **data class**
combinations. **parameter classes** should inherit from the `processParam`
class defined in {GiottoClass}

```{r, eval=FALSE}
library(methods)
library(GiottoClass)

# parameter class definition
setClass("custom_threshold", contains = "processParam")
# method definition
setMethod("processData",
    signature(x = "Matrix", param = "custom_threshold"),
    function(x, param, ...) {
        # param and ... are unused since this example function doesn't need external params
        
        # ************************************************************
        # function here should return a supported matrix-type output
        # with the same dimensions
        # ************************************************************
        
        # simple thresholding example
        x[x < 5] <- 0
        return(x)
    }
)

# run defined operation
g <- processExpression(g,
    new("custom_threshold"),
    name = "thresh5"
)

# check thresholding worked
## original
raw_mat <- g[["expression", "raw"]][[1]][]
sum(raw_mat > 0 & raw_mat < 5) # 134986

## thresholded
thresh_mat <- g[["expression", "thresh5"]][[1]][]
sum(thresh_mat > 0 & thresh_mat < 5) # 0
```


# Session Info
```{r, eval=FALSE}
sessionInfo()
```

```
R version 4.4.1 (2024-06-14)
Platform: aarch64-apple-darwin20
Running under: macOS 15.3.1

Matrix products: default
BLAS:   /System/Library/Frameworks/Accelerate.framework/Versions/A/Frameworks/vecLib.framework/Versions/A/libBLAS.dylib 
LAPACK: /Library/Frameworks/R.framework/Versions/4.4-arm64/Resources/lib/libRlapack.dylib;  LAPACK version 3.12.0

locale:
[1] en_US.UTF-8/en_US.UTF-8/en_US.UTF-8/C/en_US.UTF-8/en_US.UTF-8

time zone: America/New_York
tzcode source: internal

attached base packages:
[1] stats     graphics  grDevices utils     datasets  methods   base     

other attached packages:
[1] Giotto_4.2.1      GiottoClass_0.4.7

loaded via a namespace (and not attached):
 [1] tidyselect_1.2.1            viridisLite_0.4.2           dplyr_1.1.4                
 [4] GiottoVisuals_0.2.12        fastmap_1.2.0               SingleCellExperiment_1.26.0
 [7] lazyeval_0.2.2              digest_0.6.37               dbMatrix_0.0.0.9023        
[10] lifecycle_1.0.4             terra_1.8-29                magrittr_2.0.3             
[13] compiler_4.4.1              rlang_1.1.4                 tools_4.4.1                
[16] yaml_2.3.10                 igraph_2.1.1                utf8_1.2.4                 
[19] data.table_1.16.2           knitr_1.49                  S4Arrays_1.4.0             
[22] htmlwidgets_1.6.4           reticulate_1.39.0           DelayedArray_0.30.0        
[25] abind_1.4-8                 withr_3.0.2                 purrr_1.0.2                
[28] BiocGenerics_0.50.0         grid_4.4.1                  stats4_4.4.1               
[31] fansi_1.0.6                 colorspace_2.1-1            ggplot2_3.5.1              
[34] scales_1.3.0                gtools_3.9.5                SummarizedExperiment_1.34.0
[37] cli_3.6.3                   rmarkdown_2.29              crayon_1.5.3               
[40] generics_0.1.3              rstudioapi_0.16.0           httr_1.4.7                 
[43] rjson_0.2.21                zlibbioc_1.50.0             parallel_4.4.1             
[46] XVector_0.44.0              matrixStats_1.4.1           vctrs_0.6.5                
[49] Matrix_1.7-0                jsonlite_1.8.9              GiottoData_0.2.15          
[52] IRanges_2.38.0              S4Vectors_0.42.0            ggrepel_0.9.6              
[55] scattermore_1.2             magick_2.8.5                GiottoUtils_0.2.4          
[58] plotly_4.10.4               tidyr_1.3.1                 glue_1.8.0                 
[61] codetools_0.2-20            cowplot_1.1.3               gtable_0.3.6               
[64] GenomeInfoDb_1.40.0         GenomicRanges_1.56.0        UCSC.utils_1.0.0           
[67] munsell_0.5.1               tibble_3.2.1                pillar_1.9.0               
[70] htmltools_0.5.8.1           GenomeInfoDbData_1.2.12     R6_2.5.1                   
[73] evaluate_1.0.1              lattice_0.22-6              Biobase_2.64.0             
[76] png_0.1-8                   backports_1.5.0             SpatialExperiment_1.14.0   
[79] Rcpp_1.0.13-1               SparseArray_1.4.1           checkmate_2.3.2            
[82] colorRamp2_0.1.0            xfun_0.49                   MatrixGenerics_1.16.0      
[85] pkgconfig_2.0.3  
```
