---
title: "Giotto Object Creation - Piecewise"
output: 
  html_document:
    number_sections: true
    toc: true
pkgdown:
  as_is: true
vignette: >
  %\VignetteIndexEntry{Giotto Object Creation - Piecewise}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, eval=FALSE}
# Ensure Giotto Suite is installed.
if(!"Giotto" %in% installed.packages()) {
  pak::pkg_install("drieslab/Giotto")
}
# Ensure the Python environment for Giotto has been installed.
genv_exists <- Giotto::checkGiottoEnvironment()
if(!genv_exists){
  # The following command need only be run once to install the Giotto environment.
  Giotto::installGiottoEnvironment()
}

# Ensure Giotto Suite is installed.
if(!"GiottoData" %in% installed.packages()) {
  pak::pkg_install("drieslab/GiottoData")
}

library(Giotto)
```


Piecewise creation is most convenient when assembling a fully custom `giotto` analysis object. 
This method requires first loading the data in as Giotto-compatible subobjects, 
after which they can be directly appended to an empty `giotto` object.

For a more complete introduction to the data and Giotto subobjects used here, 
see the point data Giotto object creation [tutorial](obj_create_pnts.html) 


# Abbreviated data loading steps

```{r, eval=FALSE}
# function to get a filepath from GiottoData
mini_viz_raw <- function(x) {
  system.file(
    package = "GiottoData",
    file.path("Mini_datasets", "Vizgen", "Raw", x)
  )
}

mini_viz_poly_path <- mini_viz_raw(file.path("cell_boundaries", "z0_polygons.gz"))
mini_viz_tx_path <- mini_viz_raw("vizgen_transcripts.gz")
poly_dt <- data.table::fread(mini_viz_poly_path)
viz_gpoly <- createGiottoPolygon(poly_dt)
tx_dt <- data.table::fread(mini_viz_tx_path)
tx_dt[, global_y := -global_y] # flip values to match polys 
viz_gpoints <- createGiottoPoints(tx_dt)
```

# Create a `giotto` object through appending subobjects

```{r,eval=FALSE}
g <- giotto() # initialize empty gobject
g <- setGiotto(g, viz_gpoly)
g <- setGiotto(g, viz_gpoints)
force(g)
```

```
An object of class giotto 
>Active spat_unit:  cell 
>Active feat_type:  rna 
dimensions    : 559, 498 (features, cells)
[SUBCELLULAR INFO]
polygons      : cell 
features      : rna 
[AGGREGATE INFO]


Use objHistory() to see steps and params used
```

This is essentially the same object as the one created through `createGiottoObjectSubcellular()` as shown in the previously mentioned tutorial.

Appending additional information in this manner can be used both when objects are created from scratch using `giotto()` and when they are created through more guided approaches such as `createGiottoObject()` and technology-specific convenience functions. This method of object creation and modification can be helpful for editing the object, adding additional modalities, or setting up objects for custom workflows.


# Session Info

```{r, eval=FALSE}
sessionInfo()
```

```{r, eval=FALSE}
R version 4.4.1 (2024-06-14)
Platform: aarch64-apple-darwin20
Running under: macOS 15.0.1

Matrix products: default
BLAS:   /System/Library/Frameworks/Accelerate.framework/Versions/A/Frameworks/vecLib.framework/Versions/A/libBLAS.dylib 
LAPACK: /Library/Frameworks/R.framework/Versions/4.4-arm64/Resources/lib/libRlapack.dylib;  LAPACK version 3.12.0

locale:
[1] en_US.UTF-8/en_US.UTF-8/en_US.UTF-8/C/en_US.UTF-8/en_US.UTF-8

time zone: America/New_York
tzcode source: internal

attached base packages:
[1] stats     graphics  grDevices utils     datasets  methods   base     

other attached packages:
[1] Giotto_4.1.5      GiottoClass_0.4.3

loaded via a namespace (and not attached):
 [1] tidyselect_1.2.1            viridisLite_0.4.2           dplyr_1.1.4                
 [4] GiottoVisuals_0.2.7         R.utils_2.12.3              fastmap_1.2.0              
 [7] SingleCellExperiment_1.26.0 lazyeval_0.2.2              digest_0.6.37              
[10] lifecycle_1.0.4             terra_1.7-78                magrittr_2.0.3             
[13] compiler_4.4.1              rlang_1.1.4                 tools_4.4.1                
[16] yaml_2.3.10                 igraph_2.1.1                utf8_1.2.4                 
[19] data.table_1.16.2           knitr_1.48                  S4Arrays_1.4.0             
[22] htmlwidgets_1.6.4           reticulate_1.39.0           DelayedArray_0.30.0        
[25] abind_1.4-8                 withr_3.0.1                 purrr_1.0.2                
[28] BiocGenerics_0.50.0         R.oo_1.26.0                 grid_4.4.1                 
[31] stats4_4.4.1                fansi_1.0.6                 colorspace_2.1-1           
[34] ggplot2_3.5.1               scales_1.3.0                gtools_3.9.5               
[37] SummarizedExperiment_1.34.0 cli_3.6.3                   rmarkdown_2.28             
[40] crayon_1.5.3                generics_0.1.3              rstudioapi_0.16.0          
[43] httr_1.4.7                  rjson_0.2.21                zlibbioc_1.50.0            
[46] parallel_4.4.1              XVector_0.44.0              matrixStats_1.4.1          
[49] vctrs_0.6.5                 Matrix_1.7-0                jsonlite_1.8.9             
[52] IRanges_2.38.0              S4Vectors_0.42.0            ggrepel_0.9.6              
[55] scattermore_1.2             magick_2.8.5                GiottoUtils_0.2.1          
[58] plotly_4.10.4               tidyr_1.3.1                 glue_1.8.0                 
[61] codetools_0.2-20            cowplot_1.1.3               gtable_0.3.5               
[64] GenomeInfoDb_1.40.0         GenomicRanges_1.56.0        UCSC.utils_1.0.0           
[67] munsell_0.5.1               tibble_3.2.1                pillar_1.9.0               
[70] htmltools_0.5.8.1           GenomeInfoDbData_1.2.12     R6_2.5.1                   
[73] evaluate_1.0.0              lattice_0.22-6              Biobase_2.64.0             
[76] png_0.1-8                   R.methodsS3_1.8.2           backports_1.5.0            
[79] SpatialExperiment_1.14.0    Rcpp_1.0.13                 SparseArray_1.4.1          
[82] checkmate_2.3.2             colorRamp2_0.1.0            xfun_0.47                  
[85] MatrixGenerics_1.16.0       pkgconfig_2.0.3    
```
