---
title: "Giotto Object Creation"
css: "../assets/custom.css"
output: 
  html_document:
    number_sections: false
    toc: true
pkgdown:
  as_is: true
vignette: >
  %\VignetteIndexEntry{Giotto Object Creation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


# Spatial Omic Datatypes

Spatial omics data can be provided either as aggregated data or raw features information.

```{r, echo=FALSE}
heads <- c("Aggregated Data", "Raw Features", "Spatial Unit")
txt <- c(
    "Feature information that has been calculated for a particular set of spatial locations. <br> These are commonly presented as an expression matrix paired with a set of xy(z) coordinates in tabular or matrix format.",
    "Feature information about the tissue that has not been aggregated. These features can be paired with a spatial unit to enable aggregate data-based analyses. <br> Raw features can include molecule detections with an ID and xy(z) coordinates in tabular format (e.g., transcripts), or a matrix of intensity values, typically in image format.",
    "Spatial units provide functional context to raw features, allowing them to be aggregated and treated as units of study. These can be artificial (e.g., spatial bins), biological (e.g., cell or nuclear segmentations), or technical (e.g., arrays of spots reflecting the assay's construction). <br> Annotations can either be provided as 2D raster image/matrix-like format where the integer values of matrix cells or raster pixels both identify and define the region covered by each annotation, or as a vector-type format where the vertices of the polygons are provided."
)
imgs <- sprintf(
    '<img src="%s" style="float: right; width: 45%%;"/>',
    c(
        "images/object_creation/Giotto_Datatypes-agg.png",
        "images/object_creation/Giotto_Datatypes-ft.png",
        "images/object_creation/Giotto_Datatypes-su.png"
    )
)
knitr::asis_output(paste0(sprintf(
    '<div style="margin-bottom: 20px;">%s<div style="width: 50%%;"><h3 style="margin: 0;">%s</h3><p style="padding-left: 15px; margin-top: 5px;">%s</p></div><div style="clear: both;"></div></div>',
    imgs, heads, txt
)))
```


Raw features and spatial units can be combined to generate aggregated datasets, but the reverse transformation is generally not possible.

<br><br>


# Creating a Spatial-Omics Giotto Analysis Object

Giotto Suite can work with any of the above types of spatial omic data, and has the flexibility to handle multiple of these modalities when the data is available.
See the following for instructions on setting up a custom `giotto` object with the respective input data.


```{r, echo=FALSE}
tile_imgs <- vapply(
    c(
        "images/object_creation/1_spatplot_aggregate.png",
        "images/object_creation/4_align.png",
        "images/object_creation/9_align2.png"
    ),
    function(im_path) knitr::sew(knitr::include_graphics(im_path)),
    FUN.VALUE = character(1L)
)

tile <- sprintf(
    "<strong>%s</strong>%s", 
    c(
        "Matrix and Spatial Locations", 
        "Feature detection (points) and cell annotations (vector polygons)",
        "Staining images and cell annotations (raster mask/polygons)"
    ),
    tile_imgs
)

knitr::asis_output(paste0(
    '<div class="gallery">',
    '<a href="obj_create_agg.html" class="tile">',
        tile[[1L]],
    '</a>',
    '<a href="obj_create_pnts.html" class="tile">',
        tile[[2L]],
    '</a>',
    '<a href="obj_create_img.html" class="tile">',
        tile[[3L]],
    '</a>',
    '</div>'
))
```




# Other Datatypes and Methods

```{r, echo=FALSE}
tile_imgs <- vapply(
    c(
        "images/object_creation/other-sc.png",
        "images/object_creation/other-mod.png"
    ),
    function(im_path) knitr::sew(knitr::include_graphics(im_path)),
    FUN.VALUE = character(1L)
)

tile <- sprintf(
    "<strong>%s</strong>%s", 
    c("Single Cell Datasets", "Generic Piecewise Giotto Object Assembly"), 
    tile_imgs
)

knitr::asis_output(paste0(
    '<div class="gallery-small"">',
    '<a href="obj_create_sc.html" class="tile">',
        tile[[1L]],
    '</a>',
    '<a href="obj_create_mod.html" class="tile">',
        tile[[2L]],
    '</a>',
    '</div>'
))
```








# Convenience functions for specific platforms

There are also several convenience functions we provide for loading in data from popular platforms. These functions take care of reading the expected output folder structures, auto-detecting where needed data items are, formatting items for ingestion, then object creation. These technologies also have expanded tutorials under the examples tab.

```
createGiottoVisiumObject()
createGiottoVisiumHDObject()
createGiottoXeniumObject()
createGiottoCosMxObject()
```


# Customizing the Giotto Object

By providing values to other `createGiottoObject()` or `createGiottoObjectSubcellular()` parameters, it is possible to add:

- **Cell** or **feature (gene) metadata**: see
    [addCellMetadata](../reference/addCellMetadata.html) and
    [addFeatMetadata](../reference/addFeatMetadata.html)
- **Spatial networks** or **grids**: see [Visualizations](./visualizations.html)
- **Dimension reduction**: see [Clustering](./dimension_reduction.html)
- **Images**: see [createGiottoLargeImage](../reference/createGiottoLargeImage.html)
- **giottoInstructions**: see [Create and change Giotto instructions](instructions.html) 

For a more in-depth look at the `giotto` object structure, take a look at the [introduction to giotto classes](structure.html)

