---
title: "Xenium FFPE Human Lung Cancer with Multimodal Cell Segmentation"
output: 
  html_document:
    number_sections: true
    toc: true
pkgdown:
  as_is: true
vignette: >
  %\VignetteIndexEntry{Xenium FFPE Human Lung Cancer with Multimodal Cell Segmentation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Install Extra Packages

## _arrow_ installation

Xenium datasets requires arrow with ZTSD support to be installed to work with 
parquet files.
This is optional if you want to use a different format for the tabular data.
Building with ZTSD support may take a while.
```{r, eval=FALSE}
has_arrow <- requireNamespace("arrow", quietly = TRUE)
zstd <- TRUE
if (has_arrow) {
    # check arrow_info() to see that zstd support should be TRUE
    # See https://arrow.apache.org/docs/r/articles/install.html for details
    zstd <- arrow::arrow_info()$capabilities[["zstd"]]
}
if (!has_arrow || !zstd) {
    # install with compression library needed for 10x parquet files
    # this may take a while
    Sys.setenv(ARROW_WITH_ZSTD = "ON")
    install.packages("arrow", repos = c("https://apache.r-universe.dev"), type = "source")
}
```

## _tifffile_ and _imagecodecs_ installation

_tifffile_ is a python package for working with tif images. 
_imagecodecs_ provides the needed JPEG2000 compression codec.
10x provides their images as `ome.tif` but _Giotto_ usually either cannot open 
or incurs a large speed penalty when accessing these files. Instead we use
_tifffile_ to convert these images into normal `tif` images.

Check that _Giotto_ can access a python env and install one if it can't. Then
activate that python env.

```{r, eval=FALSE}
library(Giotto)

# Ensure Giotto can access a python env
genv_exists <- checkGiottoEnvironment()
if(!genv_exists){
  # The following command need only be run once to 
  # install a default Giotto environment
  installGiottoEnvironment()
}

# set specific python path to use or python environment name
# leaving NULL will use default (see ?set_giotto_python_path())
python_path <- NULL
set_giotto_python_path(python_path = python_path)
```

Check for presence of _tifffile_ and _imagecodecs_ in selected python env then
install if missing.
```{r, eval=FALSE}
# install packages if not already installed in environment
need_py_inst <- try(GiottoUtils::package_check(
    pkg_name = c("tifffile", "imagecodecs"), repository = c("pip:tifffile", "pip:imagecodecs")
), silent = TRUE)
if (!isTRUE(need_py_inst)) {
    active_env <- GiottoUtils::py_active_env()
    reticulate::conda_install(
        envname = active_env, packages = c("tifffile", "imagecodecs"), pip = TRUE
    )
    # may need a session restart after installation
}
```


# Dataset Explanation

This is the [10X Xenium FFPE Human Lung Cancer dataset](https://www.10xgenomics.com/datasets/preview-data-ffpe-human-lung-cancer-with-xenium-multimodal-cell-segmentation-1-standard). 
Xenium captures individual transcript detections with a spatial resolution of ~100nm, providing an extremely highly resolved subcellular spatial dataset.
This particular dataset also showcases their multimodal cell segmentation.

The Xenium Human Multi-Tissue and Cancer Panel (377) genes was used. The exported data is from their Xenium Onboard Analysis v2.0.0 pipeline.

The relevant items are:

- Xenium Output Bundle (full)
- Supplemental: Post-Xenium H&E image (OME-TIFF)
- Supplemental: H&E Image Alignment File (CSV)

## Expected Directory Structure

When unzipped, you should have the following directory structure:

<details><summary>expand</summary>
```
    ├── analysis.tar.gz
    ├── analysis.zarr.zip
    ├── analysis_summary.html
    ├── aux_outputs.tar.gz
    ├── transcripts.csv.gz
    ├── transcripts.parquet
    ├── transcripts.zarr.zip
    ├── cell_boundaries.csv.gz
    ├── cell_boundaries.parquet
    ├── nucleus_boundaries.csv.gz
    ├── nucleus_boundaries.parquet
    ├── cell_feature_matrix.tar.gz
    ├── cell_feature_matrix
    │   ├── barcodes.tsv.gz
    │   ├── features.tsv.gz
    │   └── matrix.mtx.gz
    ├── cell_feature_matrix.h5
    ├── cell_feature_matrix.zarr.zip
    ├── cells.csv.gz
    ├── cells.parquet
    ├── cells.zarr.zip
    ├── experiment.xenium
    ├── gene_panel.json
    ├── metrics_summary.csv
    ├── morphology.ome.tif
    ├── morphology_focus
    │   ├── morphology_focus_0000.ome.tif
    │   ├── morphology_focus_0001.ome.tif
    │   ├── morphology_focus_0002.ome.tif
    │   ├── morphology_focus_0003.ome.tif
    ├── Xenium_V1_humanLung_Cancer_FFPE_he_image.ome.tif
    └── Xenium_V1_humanLung_Cancer_FFPE_he_imagealignment.csv
```
</details>

<br>

The above directory structuring and naming is characteristic of Xenium v2.0 
pipeline outputs. The only items that may not be exactly the same across all 
outputs are the morphology focus directory if multimodal segmentation was not
done, and the naming of the aligned image items 
(`Xenium_V1_humanLung_Cancer_FFPE_he_image.ome.tif` and 
`Xenium_V1_humanLung_Cancer_FFPE_he_imagealignment.csv`).

For the morphology focus images, you may have fewer images if the experiment 
did not include the multimodal cell segmentation. As for the aligned images, 
this is usually done after the Xenium experiment concludes and is added on using
Xenium Explorer. Naming and location of the aligned image(s) and associated 
alignment info are entirely up to creator of the dataset.

<details>
<summary>more about xenium pre-release image types</summary>

<hr>

This dataset provides several images

**images to load**

- `Xenium_FFPE_Human_Breast_Cancer_Rep1_he_image.ome.tif`
  - post-Xenium H&E. Used with alignment matrix.
- `Xenium_FFPE_Human_Breast_Cancer_Rep1_if_image.ome.tif`
  - post-Xenium IF. Used with alignment matrix.
- `morphology_focus.ome.tif`
  - DAPI image combined from the most in-focus regions from multiple z stacks

**images not loaded**

- `morphology.ome.tif`
  - z-stacked image. Many planes may be less focused
- `morphology_mip.ome.tif`
  - [maximum intensity projection](https://en.wikipedia.org/wiki/Maximum_intensity_projection) 
  image created from the `morphology.ome.tif`.
  Has more even brightness compared against `morphology_focus.ome.tif`, but is
  also slightly blurrier. This image type was only provided with the
  pre-release and does not show up in later Xenium pipeline versions.

</details>

<br>
